---
title:  "BaseR Comments"     
author: "Christopher Witt"
date: "September 22, 2015"
output: html_document
---

### Base R Version

#### Data
Let us begin by simulating our sample data of 3 factor variables and 4 numeric variables.

```{r echo = T}
## Simulate some data

## 3 Factor Variables
FacVar1=as.factor(rep(c("level1","level2"),25))
FacVar2=as.factor(rep(c("levelA","levelB","levelC"),17)[-51])
FacVar3=as.factor(rep(c("levelI","levelII","levelIII","levelIV"),13)[-c(51:52)])

## 4 Numeric Vars
set.seed(123)
NumVar1=round(rnorm(n=50,mean=1000,sd=50),digits=2) ## Normal distribution
set.seed(123)
NumVar2=round(runif(n=50,min=500,max=1500),digits=2) ## Uniform distribution
set.seed(123)
NumVar3=round(rexp(n=50,rate=.001)) ## Exponential distribution
NumVar4=2001:2050

simData=data.frame(FacVar1,FacVar2,FacVar3,NumVar1,NumVar2,NumVar3,NumVar4)
```

<!--This is all the data for the graph and it sets the minimums, maximums, and averages of the graph. It provides a point of origin for all graphs to reference to. -->

#### One Variable: Numeric Variable

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
plot(simData$NumVar1,type="o") ## Index plot
hist(simData$NumVar1) ## histogram
plot(density(simData$NumVar1)) ## Kernel density plot
boxplot(simData$NumVar1) ## box plot
```

<!--The first line is a reference to the data given earlier. By using the type of plot chosen the graph is made within those specifications. Through choosing a different specification the data can then be used to plot a different graph with ease. This is shown when the index plot is simply changed to a histogram graph. For the third graph there is more of a trend to the graph that allows for it to be analyzed differently than its predeccesor graphs. By putting different code before the data set the graph is able to form a box plot with the variables of the original data set. -->

#### One Variable: Factor Variable

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
plot(simData$FacVar3) ## bar plot

## pie chart - Not the best graph --- use with caution
counts=table(simData$FacVar3) ## get counts
labs=paste(simData$FacVar3,counts)## create labels
pie(counts,labels=labs) ## plot
```

<!--The bar plot charts the data out in a format that makes it seem larger than the actual axises. By putting "counts" the data is then counted and by putting labs it is able to add labels to the data and then identifying it as a pie chart and ending it with plot the data is able to be turned into a pie chart.-->

#### Two Variables: Two Numeric Variables
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
plot(simData$NumVar1,type="o",ylim=c(0,max(simData$NumVar1,simData$NumVar2)))## index plot with one variable
lines(simData$NumVar2,type="o",lty=2,col="red")## add another variable

## Let's draw density plots : https://stat.ethz.ch/pipermail/r-help/2006-August/111865.html
dv1=density(simData$NumVar1) 
dv2=density(simData$NumVar2)
plot(range(dv1$x, dv2$x),range(dv1$y, dv2$y), type = "n", xlab = "NumVar1(red) and NumVar2 (blue)",
     ylab = "Density")
lines(dv1, col = "red")
lines(dv2, col = "blue")

## scatterplots
plot(simData$NumVar1,simData$NumVar2)
```

<!--The index plot adds more data to the set by including the Numvar2 within the code. In addition to this the new data is identified drawn as red which makes it easier to compare and contrast the two data sets. For the density plot the two variables have to be plotted on the same range so they both can fit within the graph. Then they are identified by their unique colors which are in this case red and blue. This color distniction allows for the graphs to be easily comparable. For the scatterplot it would have been helpful for the data to change the colors that way it would have been easier to identify them.  -->

#### Two Variables: Two Factor Variables
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## Mosaic plot
plot(table(simData$FacVar2,simData$FacVar3))

## barplots
bartable=table(simData$FacVar2,simData$FacVar3) ## get the cross tab
barplot(bartable,beside=TRUE, legend=levels(unique(simData$FacVar2))) ## plot 
barplot(bartable, legend=levels(unique(simData$FacVar2))) ## stacked
barplot(prop.table(bartable,2)*100, legend=levels(unique(simData$FacVar2))) ## stacked 100%
```

<!--Adding color to the Mosaic plot will help it show the relationship between the variables. When it is gray as it is then it makes deciphering the data all that much harder for the audience. For the 1st bar plot the color gradients help discern the differences between each level. The second bar chart is altered so that the data is stacked one on top of the other. The legend for the stacked graph is within the data itself and is hard to read. On the third stacked chart by enlargening and extending the axis the data is more easily read from the 0 to 100 scale.    -->

#### Two Variables: One Factor and One Numeric
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## Box plots for the numeric var over the levels of the factor var
plot(simData$FacVar1,simData$NumVar1) 

## density plot of numeric var across multiple levels of the factor var
level1=simData[simData$FacVar1=="level1",] 
level2=simData[simData$FacVar1=="level2",]

dv3=density(level1$NumVar1)
dv4=density(level2$NumVar1)

plot(range(dv3$x, dv4$x),range(dv3$y, dv4$y), type = "n", xlab = "NumVar1 at Level1 (red) and NumVar1 at Level2 (blue)",ylab = "Density")
lines(dv3, col = "red")
lines(dv4, col = "blue")

## Mean of one numeric var over levels of one factor var
meanagg=aggregate(simData$NumVar1, list(simData$FacVar3), mean)

dotchart(meanagg$x,labels=meanagg$Group.1) ## Dot Chart 
barplot(meanagg$x,names.arg=meanagg$Group.1)## Bar plot 
## Question: Is a bar plot even appropriate when displaying a mean--- a point?
```

<!--This chart allows for the audience to be able to interpret the differences between the two box charts by using two different data sets. For the density plot the graph axises are extended in order to have all the data within the graph. The dot chart shows the relationship of the mean for different levels. The dot chart have axises that should be more precise in its numbering because it makes it difficult to get a data sets true values. For the bar chart the axis should be extended in order to compensate for level 3 and 4 which seem to both be over the range of 1000.  -->

#### Three Variables: Three Factor Variables

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
par(mfrow=c(1,2))

bar1table=table(level1$FacVar2,level1$FacVar3)
barplot(bar1table,beside=TRUE, main="FacVar1=level1")

bar2table=table(level2$FacVar2,level2$FacVar3)
barplot(bar2table,beside=TRUE, main="FacVar1=level2", legend=levels(unique(level2$FacVar2)))
```

<!--The bartables should have made the names different from the levels that are labeling the X axis. The levels also should not have been split up as they were because it can confuse the data for the audience. It probably would have been more effective for the audience if the two data sets were combined together and would therefore allow for the audience to draw easier conclusions about the data. -->

#### Three Variables: One Numeric and Two Factor Variables
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
par(mfrow=c(1,1))
## boxplot of NumVar1 over an interaction of 6 levels of the combination of FacVar1 and FacVar2
boxplot(NumVar1~interaction(FacVar1,FacVar2),data=simData)

## Mean of 1 Numeric over levels of two factor vars
meanaggg=aggregate(simData$NumVar1, list(simData$FacVar1,simData$FacVar2), mean)
meanaggg=meanaggg[order(meanaggg$Group.1),]
meanaggg$color[meanaggg$Group.2=="levelA"] = "red"
meanaggg$color[meanaggg$Group.2=="levelB"] = "blue"
meanaggg$color[meanaggg$Group.2=="levelC"] = "darkgreen"  

dotchart(meanaggg$x,labels=meanaggg$Group.2, groups=meanaggg$Group.1,color=meanaggg$color) ## dotchart

interaction.plot(meanaggg$Group.2,meanaggg$Group.1,meanaggg$x,type="b", col=c(1:2),pch=c(18,24)) ## interaction plot - line plots of means

## some a bar plot
par(mfrow=c(1,2))

level1=meanaggg[meanaggg$Group.1=="level1",]
level2=meanaggg[meanaggg$Group.1=="level2",]

barplot(level1$x,names.arg=level1$Group.2, main="FacVar1=level1")
barplot(level2$x,names.arg=level2$Group.2, main="FacVar1=level2")
```

<!--This boxplot includes the two factors. The good thing about this boxplot is that you can see the median trends well as well as the upper quartile trends. The only problem with this graph is that the X axis needs to be labeled better so that the audience can tell which box is which. This dot chart looks at the mean and separates the two sets of data into two different levels of graphs. This is fairly easy to read because the data points are easily identifiable. For the interaction plot the graph is looking at the mean within levels of level 1 and 2. One problem with this graph is that it could be labeled better. For the bar plot the range should be increased so that the Y axis can accomodate both graphs. Changing the color would be good as well because it would add more variety to the graph. -->

#### Three Variables: Two Numeric and One Factor Variables

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## Scatter plot with color identifying the factor variable
par(mfrow=c(1,1))
plot(simData$NumVar1,simData$NumVar2, col=simData$FacVar1)
legend("topright",levels(simData$FacVar1),fill=simData$FacVar1)
```

<!--For the scatter plot the data should try and be placed in increments of 100 that would make the data more uniform on the X and Y axis in terms of the graph. The legend for the level 1 and 2 should be moved out of the range of the graph because they make the audience wonder whether or not there is data hidden behind the legend.  -->

#### Three Variables: Three Numeric Variables 

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## NumVar4 is 2001 through 2050... possibly, a time variable - use that as the x-axis
plot(simData$NumVar4,simData$NumVar1,type="o",ylim=c(0,max(simData$NumVar1,simData$NumVar2)))## join dots with lines

lines(simData$NumVar4,simData$NumVar2,type="o",lty=2,col="red")## add another line

## Bubble plot - scatter plot of NumVar1 and NumVar2 with individual observations sized by NumVar3
# http://flowingdata.com/2010/11/23/how-to-make-bubble-charts/
radius <- sqrt( simData$NumVar3/ pi )
symbols(simData$NumVar1,simData$NumVar2,circles=radius, inches=.25,fg="white", bg="red", main="Sized by NumVar3")
```

<!--This graph feels like there is too much information trying to squeeze into one graph. Maybe breaking up the graph into different segments, like say by each decade, would ease the eye sore that this graph gives me. In addition to this the graph lines are also hard to follow and should be changed to something more solid that is easier to register change with. The bubble plot is hard to determine the radius of the different variables. In addition to this not all of the shapes are the same within this graph.  -->

#### Scatterplot Matrix of all Numeric Vars, colored by a Factor variable
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
pairs(simData[,4:7], col=simData$FacVar1)
```

<!--This is too difficult to decipher for the average audience. Making it one combined axis would probably make the data easier to view.  -->

#### References

Besides the link from flowingdata.com referred to in the context of the bubble plot, additional websites were used as references. 
http://www.harding.edu/fmccown/r/
http://www.statmethods.net/
